---
title: "13: High Frequency Data"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Determine stormflow and baseflow from high frequency flow data
2. Use a hysteresis plot to understand watershed dynamics
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

High frequency data is usually defined as frequencies significantly lower than daily (e.g. 5-minute, 15-minute, 1 hr etc). What types of hydrological and biological processes happen on this timescale that we might want to investigate?

<<<<<<< HEAD
> temp/oxygen -->big daily; fast storms/flash floods; contaminants
=======
>
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771


## Session Set Up

```{r, message = FALSE}
getwd()

<<<<<<< HEAD
install.packages("EcoHydRology")

=======
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771
packages <- c(
  "dataRetrieval", "tidyverse", "cowplot", "EcoHydRology", "xts", "dygraphs"
  )
invisible(lapply(packages, library, character.only = TRUE)) 

theme_set(theme_classic(base_size = 12))
```

## Pull Data

```{r download data}

dischargedatTFC2018 <- readNWISuv(
  site = "02097280", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01")%>%
  #) %>% 
#<<<<<<< HEAD
  renameNWISColumns() # since the datetime has hours, notice there is a time zone code, which is ts_cd; UTC time corrodinates
#=======
  renameNWISColumns()
#>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771


dischargedatEC2018 <- readNWISuv(
  site = "0208675010", #Ellerbe Creek at Club Boulevard at Durham, NC
  parameterCd = "00060", 
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()


plot_grid(
  ggplot(dischargedatTFC2018, aes(x = dateTime, y = Flow_Inst)) +
    geom_line() + 
    scale_y_log10() +
    ggtitle("Third Fork Creek"),
  ggplot(dischargedatEC2018, aes(x = dateTime, y = Flow_Inst)) + 
    geom_line() + 
    scale_y_log10() +
    ggtitle("Ellerbe Creek"),
  ncol = 1
)
          

```

The shape and size of river hydrographs following rain events can tell us how water moves through a watershed. Once rain falls, how does water get to a stream? What would the streamflow do?

<<<<<<< HEAD
> overland flow; groundwater; fall in the river channel; pipes; normal
> in Ellerbe Creek, the base flow is in the just a little higher above the base line; quickflow
=======
>
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771


_"Hydrograph separation is one of the most desperate analysis techniques in use in hydrology" - Hewlett & Hibbert 1967_


```{r baseflowseparation}
TFCbaseflow <- EcoHydRology::BaseflowSeparation(
  dischargedatTFC2018$Flow_Inst, 
  filter_parameter = 0.925, 
  passes = 3
  )

TFC2018 <- cbind(dischargedatTFC2018, TFCbaseflow)

ggplot(TFC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
<<<<<<< HEAD
   #scale_y_log10() +
=======
  # scale_y_log10() +
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(TFC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(TFC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(TFC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()

Export <- TFC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
         baseflowexport = bt * timestep,
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)
<<<<<<< HEAD

baseflow_percen = 
=======
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771
  
```

What percentage of water is exported from this watershed as baseflow? What percentage as quickflow?

<<<<<<< HEAD
> quickflow arrives the peak earlier, thus the early part of the flow would be quickflow; base flow has to go through the soil porcess, it would pass through later. When a storm happens, the quick flow is the first contribution, and serveral hours passed, base flow becomes the new contribution. we have more overland flow, which has impermeable soil surface, which means we have more quickflow than baseflow.

> For TFC: 51.9% baseflow and 48.1% quickflow; for ECL 43.3% baseflow and 56.6% quickflow.

Repeat this analysis for Ellerbe Creek. How do the percentages compare?

```{r}

ECbaseflow <- BaseflowSeparation(
  dischargedatEC2018$Flow_Inst, 
  filter_parameter = 0.925, 
  passes = 3
  )

EC2018 <- cbind(dischargedatEC2018, ECbaseflow)

ggplot(EC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
   #scale_y_log10() +
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(EC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(EC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(EC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()

Export <- EC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
         baseflowexport = bt * timestep,
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)


```
=======
> 

Repeat this analysis for Ellerbe Creek. How do the percentages compare?

>
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771

What does this mean physically about the watershed?

>

## Hydrologic Flashiness

<<<<<<< HEAD
A "flashy" stream increases in flow faster following a precipitation event. a fast, dramatic change. quickflow 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> increase

What do you think "urban stream syndrom" refers to?

>hydrologic flashiniess is a syptom of "urban stream syndrom", urban stream tend to be deeper

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as $$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$  j is time step 1,
small catchment area can generate more flassiniess.
=======
A "flashy" stream increases in flow faster following a precipitation event. 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> 

What do you think "urban stream syndrom" refers to?

>

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as $$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771

```{r Flashiness}
TFCsitedata <- readNWISsite(site = "02097280")
TFC.catchment.size <- TFCsitedata$drain_area_va #square miles

RBI_TFC2018 <- (
  sum(abs(diff(dischargedatTFC2018$Flow_Inst))) / 
    sum(dischargedatTFC2018$Flow_Inst[-1])
  ) / TFC.catchment.size

ECsitedata <- readNWISsite(site = "0208675010")
EC.catchment.size <- ECsitedata$drain_area_va #square miles

RBI_EC2018 <- (
  sum(abs(diff(dischargedatEC2018$Flow_Inst))) / 
    sum(dischargedatEC2018$Flow_Inst[-1])
  ) / EC.catchment.size

```

Which stream is flashier?

<<<<<<< HEAD
>EC is flashier, RBI_EC >RBI_TFC

```{r}
dischargedatHR2018 <- readNWISuv(
  site = "03449500", 
  parameterCd = "00060",
  startDate = "", 
  endDate = ""
  ) %>% 
  renameNWISColumns()
```

=======
>
>>>>>>> a41323e1b2a0e73e03c36d497f51177888a2e771

## Chemical Flashiness

Chemical concentrations in rivers can be either highly variable with discharge, or "chemostatic", which means the concentration barely changes with discharge. 

```{r}
CCdat <- readNWISuv(
  site = "02249500", # CRANE CREEK AT MELBOURNE, FL
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2015-01-1",
  endDate = "2016-01-01"
  ) %>%
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(CCdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


Potdat <- readNWISuv(
  site = "01646500", # POTOMAC RIVER NEAR WASH, DC LITTLE FALLS PUMP STA
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2016-01-1",
  endDate = "2017-01-01"
  ) %>% 
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(Potdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Which of these rivers is more chemostatic?

>

## Hysteresis Loops

Hysteresis occurs when the concentration varies with discharge in a different relationship on the rising and falling limbs of a storm. This can be caused by _flushing_ behavior or _dilution_ behavior. 

```{r}
dygraph(
  cbind(
    Flow = xts(Potdat$Flow_Inst, order.by = Potdat$dateTime), 
    Nitrate = xts(Potdat$Nitrate_mgl, order.by = Potdat$dateTime)
    )
  ) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector()
```

Look at the storm that occured around February 1st. Do you expect this storm to be a flushing or a diluting storm?

>diluting storm

```{r}
PotomacStorm <- Potdat %>%
  filter(dateTime > "2016-01-31" & dateTime < "2016-02-14") 

ggplot(PotomacStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() 


```

Did this storm exhibit clockwise or counterclockwise hysteresis?

> counterclockwise 

Which direction do hysteresis loops go for diluting storms?

>decrease first then increase

Which direction do hysteresis loops go for flushing storms?

>increase first then decrease

## Closing Discussion

What hydrological processes did we investigate today? 

>

What can high-frequency data be used for that daily data may not be able to capture?

> 
