---
title: "Assignment 5: Water Quality in Lakes"
author: "Yixin Wen"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in lakes

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A05_Salk.html") prior to submission.

The completed exercise is due on 2 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and LAGOSNE packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the LAGOSdata database and the trophic state index csv file we created on 2019/09/27.

```{r setup}
getwd()

library(tidyverse)
library(lubridate)
library(LAGOSNE)

theme_set(theme_classic())
load(file = "./Data/Raw/LAGOSdata.rda")
#load(file = "./Data/LAGOStrophic.csv") this cannot work, why?
LAGOStrophic <-read_csv("./Data/LAGOStrophic.csv")
```

## Trophic State Index

5. Similar to the trophic.class column we created in class (determined from TSI.chl values), create two additional columns in the data frame that determine trophic class from TSI.secchi and TSI.tp (call these trophic.class.secchi and trophic.class.tp). 

```{r}
LAGOStrophic <-
  mutate(LAGOStrophic,
         trophic.class.secchi =
           ifelse(TSI.secchi < 40, "Oligotrophic", 
                   ifelse(TSI.secchi < 50, "Mesotrophic",
                          ifelse(TSI.secchi < 70, "Eutrophic", "Hypereutrophic"))),
         trophic.class.tp = 
           ifelse(TSI.tp < 40, "Oligotrophic",
                  ifelse(TSI.tp < 50, "Mesotrophic",
                         ifelse(TSI.tp < 70, "Eutrophic", "Hypereutrophic")))) 
  
```


6. How many observations fall into the four trophic state categories for the three metrics (trophic.class, trophic.class.secchi, trophic.class.tp)? Hint: `count` function.

```{r}
trophic.class <-
  LAGOStrophic %>%
    count(trophic.class)
unique(trophic.class)

trophic.class.secchi <-
  LAGOStrophic %>%
    count(trophic.class.secchi)
unique(trophic.class.secchi)

trophic.class.tp <-
  LAGOStrophic %>%
    count(trophic.class.tp)
unique(trophic.class.tp)
```

> for trophic.class, there are 41861 in Eutrophic, 14379 in Hpereutrophic, 15413 in Mesotrophic and 3298 in Oligotrophic; for trophic.class.secchi, there are 28659 in Eutrophic, 5099 in Hpereutrophic, 25083 in Mesotrophic and 16110 in Oligotrophic; for trophic.class.tp, there are 24839 in Eutrophic, 7228 in Hpereutrophic, 23023 in Mesotrophic and 19861 in Oligotrophic.

7. What proportion of total observations are considered eutrohic or hypereutrophic according to the three different metrics (trophic.class, trophic.class.secchi, trophic.class.tp)?

```{r}
trophic.class.pro = (trophic.class$n[1]+trophic.class$n[2])/74951
trophic.class.secchi.pro = (trophic.class.secchi$n[1]+trophic.class.secchi$n[2])/74951
trophic.class.tp.pro = (trophic.class.tp$n[1]+trophic.class.tp$n[2])/74951
```

> In trophic.class, the proportion is 0.75; in trophic.class.secchi, the proportion is 0.45;
in trophic.class.tp, the proportion is 0.43.

Which of these metrics is most conservative in its designation of eutrophic conditions? Why might this be?

> trophic.class.tp is the most conservative metrics in eutrophic conditions.

Note: To take this further, a researcher might determine which trophic classes are susceptible to being differently categorized by the different metrics and whether certain metrics are prone to categorizing trophic class as more or less eutrophic. This would entail more complex code.

## Nutrient Concentrations

8. Create a data frame that includes the columns lagoslakeid, sampledate, tn, tp, state, and state_name. Mutate this data frame to include sampleyear and samplemonth columns as well. Call this data frame LAGOSNandP. 

```{r}
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr

LAGOSlocus$lagoslakeid <- as.factor(LAGOSlocus$lagoslakeid)
LAGOSnutrient$lagoslakeid <- as.factor(LAGOSnutrient$lagoslakeid)

LAGOSlocations <- left_join(LAGOSlocus, LAGOSstate, by = "state_zoneid")

LAGOSNandP <- 
  left_join(LAGOSnutrient, LAGOSlocations, by = "lagoslakeid") %>%
  select(lagoslakeid, sampledate, tn, tp, 
         state, state_name) %>%
  mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate))
```

9. Create two violin plots comparing TN and TP concentrations across states. Include a 50th percentile line inside the violins.
```{r}
stateTNviolin <- ggplot(LAGOSNandP, aes(x = state, y = tn)) +
  geom_violin(draw_quantiles = 0.50)+
  labs(x = "State", y = expression(Total_N ~ a ~ (mu*g / L)))
print(stateTNviolin)

stateTPviolin <- ggplot(LAGOSNandP, aes(x = state, y = tp)) +
  geom_violin(draw_quantiles = 0.50)+
  labs(x = "State", y = expression(Total_P ~ a ~ (mu*g / L)))
print(stateTPviolin)
```

```{r}
Nstate <-
  LAGOSNandP %>%
  select(state,tn)%>%
  group_by(state)%>%
  summarise(tn0.5 = quantile(tn, probs = 0.5, na.rm = TRUE))%>%
  mutate(tn0.5)

Pstate <-
  LAGOSNandP %>%
  select(state,tp)%>%
  group_by(state)%>%
  summarise(tp0.5 = quantile(tp, probs = 0.5, na.rm = TRUE))%>%
  mutate(tp0.5)
```

Which states have the highest and lowest median concentrations?

> TN: IA has the highest median concentrations, VT and NH have the lowest median concentrations.

> TP: IL has the highest median concentrations, MI,NH and ME have the lowest
median concentrations.

Which states have the highest and lowest concentration ranges?

> TN: IA has highest concentration ranges, VT and NH have the lowest concentration ranges.

> TP: IL has highest concentration ranges, CT and PA have the lowest concentration ranges. 

10. Create two jitter plots comparing TN and TP concentrations across states, with samplemonth as the color. Choose a color palette other than the ggplot default.

```{r}
stateTNjitterbymonth <-  
ggplot(LAGOSNandP, 
       aes(x = state, y = tn, color = samplemonth)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "State", y = expression(Total_N ~ a ~ (mu*g / L)), color = "Month") +
  scale_color_viridis_c(option = "inferno")
print(stateTNjitterbymonth)

stateTPjitterbymonth <-  
ggplot(LAGOSNandP, 
       aes(x = state, y = tp, color = samplemonth)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "State", y = expression(Total_P ~ a ~ (mu*g / L)), color = "Month") +
  scale_color_viridis_c(option = "inferno")
print(stateTPjitterbymonth)

```

```{r}
Nmonth <-
  LAGOSNandP%>%
  select(state,tn)
Nmonth <- na.omit(Nmonth)

NCounts <- ggplot(Nmonth, aes(x = state)) + 
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(NCounts)

  
```

Which states have the most samples? How might this have impacted total ranges from #9?

> TN:

> TP: 

Which months are sampled most extensively? Does this differ among states?

> TN:

> TP: 

11. Create two jitter plots comparing TN and TP concentrations across states, with sampleyear as the color. Choose a color palette other than the ggplot default.

```{r}
stateTNjitterbyyear <-  
ggplot(LAGOSNandP, 
       aes(x = state, y = tn, color = sampleyear)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "State", y = expression(Total_N ~ a ~ (mu*g / L)), color = "Year") +
  scale_color_viridis_c(option = "inferno")
print(stateTNjitterbyyear)

stateTPjitterbyyear <-  
ggplot(LAGOSNandP, 
       aes(x = state, y = tp, color = sampleyear)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "State", y = expression(Total_N ~ a ~ (mu*g / L)), color = "Year") +
  scale_color_viridis_c(option = "inferno")
print(stateTPjitterbyyear)

```


Which years are sampled most extensively? Does this differ among states?

> TN:

> TP: 

## Reflection
12. What are 2-3 conclusions or summary points about lake water quality you learned through your analysis?

>

13. What data, visualizations, and/or models supported your conclusions from 12?

> 

14. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>

15.	How did the real-world data compare with your expectations from theory?

>
