---
title: "Assignment 6: Time Series Analysis"
author: "Yixin Wen"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on time series analysis

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single pdf file.
5. After Knitting, submit the completed exercise (pdf file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A06_Salk.html") prior to submission.

The completed exercise is due on 11 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, trend, and dataRetrieval packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the ClearCreekDischarge.Monthly.csv file from the processed data folder. Call this data frame ClearCreekDischarge.Monthly.

```{r setup}
getwd()

library(tidyverse)
library(lubridate)
library(trend)
library(dataRetrieval)
library(forecast)
library(tseries)

theme_set(theme_classic())

library(readr)
ClearCreekDischarge.Monthly <- read_csv("/Users/yixinwen/Box/Duke/2019 Fall/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Data/Processed/ClearCreekDischarge.Monthly.csv")
```

## Time Series Decomposition
5. Create a new data frame that includes daily mean discharge at the Eno River for all available dates (`siteNumbers = "02085070"`). Rename the columns accordingly.
6. Plot discharge over time with geom_line. Make sure axis labels are formatted appropriately.
7. Create a time series of discharge
8. Decompose the time series using the `stl` function.
9. Visualize the decomposed time series.
```{r}
# create a data frame of Eno River 
EnoRiverDischarge <- readNWISdv(siteNumbers = "02085070",
                                parameterCd = "00060",
                                startDate = "",
                                endDate = "")
names(EnoRiverDischarge)[4:5] <-c("Discharge","Approval.Code")
class(EnoRiverDischarge$Date)

# plot discarge over time
EnoRiverDischarge.plot <-
  ggplot(data = EnoRiverDischarge, aes(x = Date, y = Discharge))+
  geom_line()+
  labs(x = "Date", y = expression("Discharge (ft"^3*"/s)"))
print(EnoRiverDischarge.plot)

#create a time series of discharge
EnoRiver_ts <- ts(EnoRiverDischarge[[4]], frequency = 365)

# decompose the time series
EnoRiver_Decomposed <- stl(EnoRiver_ts, s.window = "periodic")

# visualize the decompozation
plot(EnoRiver_Decomposed)
```

10. How do the seasonal and trend components of the decomposition compare to the Clear Creek discharge dataset? Are they similar in magnitude?

> Seasonal: the range of seasonal change of Eno River discharge is smaller than that of Clear Creek discharge. The Eno River has a more compased seasonal change compared to Clear Creek.

> Trend: The range of trend component of Eno River discharge is similar to that of Clear Creek discharge.

## Trend Analysis
Research question: Has there been a monotonic trend in discharge in Clear Creek over the period of study?

11. Generate a time series of monthly discharge in Clear Creek from the ClearCreekDischarge.Monthly data frame. This time series should include just one column (discharge). 
12. Run a Seasonal Mann-Kendall test on the monthly discharge data. Inspect the overall trend and the monthly trends. 
```{r}
ClearCreek_ts <- ts(ClearCreekDischarge.Monthly[[3]], frequency = 12)

ClearCreekTrend <- smk.test(ClearCreek_ts)

ClearCreekTrend
summary(ClearCreekTrend)
```

13. Is there an overall monotonic trend in discharge over time? If so, is it positive or negative?

> since the p-value = 0.09719, which is larger than 0.05, there's no monotonic trend in discharge over time.

14. Are there any monthly monotonic trends in discharge over time? If so, during which months do they occur and are they positive or negative?

> June and July, since the z scores are larger than 0, they are positive.

## Reflection
15. What are 2-3 conclusions or summary points about time series you learned through your analysis?

> 1. when p-value is larger than 0.05, there's no significant trend on change; when p-value is smaller than 0.05, there is monotonic trend.
2. when z score is greater than 0, the trend is positive, and when z score is smaller than 0, the trend is negative. 

16. What data, visualizations, and/or models supported your conclusions from 12?

> the Seasonal Mann-Kendall test on ClearCreek can give the results of whether there is monotonic trend on discharge.

17. Did hands-on data analysis impact your learning about time series relative to a theory-based lesson? If so, how?

>Hands-on data analysis can let me explore the theory on my own, and it can help me understand it better from examples. 

18.	How did the real-world data compare with your expectations from theory?

>  The real-world data may have mistakes in the dataset. Before we do analysis on it, we need to remove the abnormal data, otherwise the results may not be correct.
